---
title: "A brief overview of CTA bus and Divvy trends"
output: pdf_document
---

```{r echo=FALSE, warning = FALSE, message=FALSE }
libs <- c("readr","haven","dplyr","tidyr","stringr","ggplot2","lubridate","reshape2","reshape")
l <- lapply(libs,library,character.only=TRUE)
```

```{r echo=FALSE, warning = FALSE, message=FALSE }
setwd("/Users/yuxiwu/git_svn/DV17/project")
bus <- read_csv("busridership_byroute.csv")
bus_rides <- bus[,!(names(bus) %in% c("daytype"))]
divvy <- read_csv("../hw2_charts/Divvy_Trips_2017_Q2.csv")

divvy <- na.omit(divvy)

bus_rides <- na.omit(bus_rides)
bus_rides$date <- str_trim(bus_rides$date)
bus_rides$date <- as.Date(strptime(bus_rides$date, format="%m/%d/%Y", tz = ""))
```

```{r echo=FALSE, warning = FALSE, message=FALSE }
yx_title <- theme(plot.title = element_text(family="CM Roman", face="bold", size=20, hjust=0, margin = margin(0, 22, 0, 0)))
yx_sub <- theme(plot.subtitle=element_text(size=11, family="CM Roman", hjust=0, face="italic", color="#8E8883", margin = margin(10, 22, 22, 0)))
yx_axis <- theme(axis.title = element_text(family = "CM Roman",size=12, face="italic",color="#635F5D", margin = margin(22, 0, 22, 0)))
axis_labels <- theme(axis.text=element_text(size=10))
yx_caption <- theme(plot.caption=element_text(family = "Helvetica", hjust = 0, size=8, face="italic", color="black"))
panel_background <- theme(panel.background = element_rect(fill = '#bdbdbd'))
color_plot <- theme(plot.background = element_rect(fill = '#E5E2E0'))
```


## General breakdown of Divvy subscribers

```{r echo=FALSE, warning = FALSE, message=FALSE }
# looking at gender counts
gen <- count(divvy, gender)
gen <- dplyr::rename(gen, "count"="n")
gen <- dplyr::rename(gen, "Gender"="gender")
gen$pct <- round(((gen$count / sum(gen$count)) * 100), digits=2)
gen$pct <-  str_pad(as.character(gen$pct), pad="%", width=6, side="right")
labels.gen <- paste(factor(gen$Gender), "\n", gen$pct, sep="")

pie_divvy_gen <- ggplot(gen, aes(x="", y=count, fill=Gender)) +
  geom_bar(stat="identity", width=1) +
  geom_text(aes(label=labels.gen), position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values=c("#ff8d1e","dodgerblue1")) +
  coord_polar(theta="y") +
  theme_void() +
  theme(legend.position = "none") + 
  labs(title="\nMale Divvy subscriber journeys outnumber female three to one",
       caption="Source: Divvy Bikes system data, Q1 & Q2 2017") +
  yx_title + yx_sub + yx_caption


pie_divvy_gen
```

## When do subscribers ride Divvy?
Again, looking only at data from the first two quarters of this year, we can already spot a possibly-season-related upward trend in ridership.  This makes sense -- as the weather gets warmer, people would more likely want to ride bikes.

```{r echo=FALSE, warning = FALSE, message=FALSE }
divvy$Date <- as.Date(strptime(divvy$start_time, format="%m/%d/%Y %H:%M:%S", tz = ""))
date_agg <- count(divvy, Date)
date_agg <- dplyr::rename(date_agg, "Journeys"="n")

divvy_line <- ggplot(date_agg, aes(x=Date, y=Journeys, group=1)) +
  geom_line(aes(color=I("dodgerblue1"))) +
  theme_minimal() +
  theme(plot.margin = unit(c(1,1,1,1), "cm")) +
  labs(title="\nDivvy Ridership increases through season changes",
       caption="\nSource: Divvy Bikes system data, Q1 & Q2 2017",
       x="Date in 2017",
       y="Journeys")
divvy_line
```


## Types and durations of Divvy journeys
A chief criticism of bikeshare systems in any major city is that they are gimmicky and tourist-oriented.  Basic arithmetic in fare calcuation, however, would demonstrate that using Divvy becomes very expensive as soon as you cross the half-hour mark, a decidedly tourist-unfriendly reality. Although, due to the nature of ridership data, it's impossible to pick out which riders are tourists, it's unlikely that those unfamiliar with the city would want to face the obstacle of finding docking stations in untested territories.  Add a 30-minute constraint on top and it's probably more likely that the majority of people who use Divvy are those who know where they're going.  Indeed, rides drop to near-zero past the 30 minute mark:

```{r echo=FALSE, warning = FALSE, message=FALSE }
#looking at types of journeys made by duration
divvy <- transform(divvy, group=cut(tripduration,
                                    breaks=c(-Inf,300, 900, 1800, 2700, 3600, Inf),
                                    labels=c('<5 minutes', '5-15 minutes', '15-30 minutes', '30-45 minutes','45-60 minutes', '60+ minutes')))
divvy <- dplyr::rename(divvy,"tripdurtype"="group")
durations <- count(divvy, tripdurtype)
durations <- dplyr::rename(durations, "Duration"="tripdurtype", "count"='n')

bar_durations <- ggplot(durations, 
              aes(x=Duration, y=count, fill=Duration)) +
  geom_bar(stat="identity",
           fill="#FF6666") +
  theme_minimal() + 
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        legend.position="none",
        plot.margin = unit(c(1,1,1,1), "cm")) +
  labs(title="\nTypes of Divvy journeys made, by duration",
       caption="\nSource: Divvy Bikes system data, Q1 & Q2 2017")

bar_durations
```


## People aren't riding buses as much as before
```{r echo=FALSE, warning = FALSE, message=FALSE }
bus_month <- bus_rides %>% group_by(month=floor_date(date, "month")) %>%
  summarize(amount=sum(rides))
bus_month <- dplyr::rename(bus_month, "Rides"="amount")
bus_month <- dplyr::rename(bus_month, "Month"="month")
bus_month <- transform(bus_month, Rides=Rides/1000000)

bus_line <- ggplot(bus_month, aes(x=Month, y=Rides, group=1)) +
  geom_line(aes(color=I("dodgerblue1"))) +
  theme_minimal() +
  theme(plot.margin = unit(c(1,1,1,1), "cm")) +
  labs(title="\nDownward trend in city-wide bus ridership since summer 2013",
       caption="\nSource: Daily Bus Ridership Totals, City of Chicago",
       x="Date",
       y="Rides (millions)")
bus_line
```


## The busiest buses reach the farthest corners of the city
```{r echo=FALSE, warning = FALSE, message=FALSE }
top_buses <- aggregate(bus_rides$rides, by=list(Category=bus_rides$route), FUN=sum)
top_buses <- dplyr::rename(top_buses, "Rides"="x")
top_buses <- dplyr::rename(top_buses, "Route"="Category")
top_buses <- transform(top_buses, Rides=Rides/1000000)

top_buses <- top_buses[order(-top_buses$Rides),][1:20,]
tb <- ggplot(top_buses, aes(x=reorder(Route, -Rides),y=Rides)) + 
  geom_bar(stat='identity') +
  theme_minimal() +
  theme(plot.margin = unit(c(1,1,1,1), "cm")) +
  labs(title="\n79th Street bus is busiest in Chicago",
       subtitle="South Side dominates in most-ridden routes\n",
       caption="\nSource: Daily Bus Ridership Totals, City of Chicago",
       x="Route",
       y="Rides (millions)")
tb
```


## Even the busiest bus is less busy now
```{r echo=FALSE, warning = FALSE, message=FALSE }
bus_rides$mo <- strftime(bus_rides$date, "%m")
bus_rides$yr <- strftime(bus_rides$date, "%Y")

b79 <- subset(bus_rides, route=="79")

b79m <- data.frame(matrix(, nrow=12, ncol=0))
y14 <- filter(bus_rides, (yr=="2014" & route=="79")) %>% 
  group_by(mo) %>%
  summarize(amount=sum(rides))
y15 <- filter(bus_rides, (yr=="2015" & route=="79")) %>%
  group_by(mo) %>%
  summarize(amount=sum(rides))
y16 <- filter(bus_rides, (yr=="2016" & route=="79")) %>% 
  group_by(mo) %>%
  summarize(amount=sum(rides))

b79m$y14 <- y14$amount / 10000
b79m$y15 <- y15$amount / 10000
b79m$y16 <- y16$amount / 10000
b79m$month <- c("01","02","03","04","05","06","07","08","09","10","11","12")

b79m_melt <- melt(b79m, id.vars="month")

bus79 <- ggplot(b79m_melt, aes(x=month, y=value, group=variable, colour=variable)) + 
  geom_line() +
  theme_minimal() +
  theme(plot.margin = unit(c(1,1,1,1), "cm")) +
  scale_color_manual(labels = c("2014", "2015","2016"), values = c("#d2e8ff", "#bbddff","#ff8d1e")) +
  labs(title="\nYearly ridership on 79 bus route",
       subtitle="Source: Daily Bus Ridership Totals, City of Chicago",
       x="Month",
       y="Rides (10,000)",
       color = "Year\n")

bus79
```

